---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# bis557

<!-- badges: start -->
<!-- badges: end -->

The goal of bis557 is to collate my function implementations, vignettes, and other files from Professor Michael Kane's Computational Statistics (BIS557) taught Fall 2019.

## Installation

You can install the development version of bis557 from [GitHub](https://github.com/WillDuke/bis557.git) with:

``` r
# install.packages("devtools")
devtools::install_github("WillDuke/bis557")
```
## Example

This package includes a variety of functions including a minimal implementation of ridge_regression:

```{r example}
library(bis557)
data("iris")
ridge_regression(Sepal.Length ~ ., iris, lambda = 0.1)
```

Vignettes in this package include answers to homework problems from the CASL book and class as well as common use-cases for some of the functions in the package.
